Profitability Prediction of Consulting Projects in the Construction Industry

### 2016-08-02 finalwriteup/draft.rmd
just conclusion left now

library('knitr')
library('ggplot2')
library("plyr")
library("dplyr")
library('magrittr')
library('reshape2')
library("rpart")
library('car')
library('e1071')
library('party')
library('randomForest')
library('RColorBrewer')
library('pwr')
library('scales')

@@

A model that is perfectly classified would have all positives correctly classified (1.0 True Positive Rate) and no incorrectly classified positives (0.0 False Positive Rate). If a model has a 1.0 TPR and 1.0 FPR (top right corner) it has correctly classified all positives at the expense of incorrectly classifying all negatives as positive. It is 'dumbly' classifying all cases as positive. If a model classifies 0.8 of its positives as TP but also 0.8 of its negative cases as positive it is 'dumb' in a similar way to the previous example. There is an 80% chance *any* case will be classified as positive. Therefore, a point that lies on the diagonal line classifies cases with the same capability as random chance.

@@

For binary classification, AUC is a more meaningful statistic than classification accuracy when the output is a probability that can be applied to the problem. The AUC indicates model performance across many thresholds while classification accuracy represents a single threshold. The AUC statistic is therefore useful in problems where cases can be ranked and a cut-off threshold that will optimise the rates of true positives and true negatives can be found [@Huang2005]. Note the 'probability' outputs from the models are not true statistical probabilities, but a score the model has assigned to a case using its own measure of certainty.

@@

ggplot(reduc.totAUC, aes(x= factor(L1), y= value)) + 
        geom_violin(aes(fill = sigdiff, colour = sigdiff)) + 
        geom_jitter(alpha = 0.2, shape = 16, colour= 'gray70') +
        stat_summary(fun.y="mean", geom="point", colour = "gray60", shape =3) +
        scale_fill_manual(values = c('sienna1','navajowhite'),
                          name = "Significantly different from boosted trees") +
        scale_colour_manual(values = c('sienna1','navajowhite'),
                          name = "Significantly Different from Boosted Trees") +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "AUC Distribution from 100 models of Each Method") +
        scale_x_discrete(labels = c("boost" = "Boosted Trees", 
                         "log" = "Logistic Regression",
                         "naiive" = "Naive Bayes", "rf" = "Random Forest")) +
        theme(legend.position = "bottom")

@@

ggplot(mice.totAUC, aes(x= factor(L1), y= value)) + 
        geom_violin(aes(fill = sigdiff, colour = sigdiff)) + 
        geom_jitter(alpha = 0.2, shape = 16, colour= 'gray70') +
        stat_summary(fun.y="mean", geom="point", colour = "gray60", shape =3) +
        scale_fill_manual(values = c('sienna1','navajowhite'),
                          name = "Significantly different from boosted trees") +
        scale_colour_manual(values = c('sienna1','navajowhite'),
                          name = "Significantly different from boosted trees") +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "AUC Distribution from 100 Models of Each Method - Imputed Data") +
        scale_x_discrete(labels = c("boost" = "Boosted Trees", 
                         "log" = "Logistic Regression",
                         "naiive" = "Naive Bayes", "rf" = "Random Forest")) +
        theme(legend.position = "bottom")
```

@@

ggplot(combine.micer, aes(x= factor(L1), y= value)) +
        geom_violin(aes(fill = Data_Source, colour = Data_Source)) +
        geom_jitter(alpha = 0.2, shape = 16, colour= 'gray70') +

        # stat_summary(fun.y="mean", geom="point", colour = "gray60", shape =3) +
        scale_fill_manual(values = c('sienna1','navajowhite'),
                          name = "Data Source") +
        scale_colour_manual(values = c('sienna1','navajowhite'),
                          name = "Data Source") +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "Comparison of Original Data vs. Imputed Data") +
        scale_x_discrete(labels = c("boost" = "Boosted Trees", 
                         "log" = "Logistic Regression",
                         "naiive" = "Naive Bayes", "rf" = "Random Forest")) +
        theme(legend.position = "bottom")

@@

Boosted trees, Random Forests and Logistic Regression all had mean AUC values around 0.775. The tests were repeated using imputed data and again the means sat around 0.775. The variables were not particularly well suited for imputing so this result is logical and could have even diminished the mean AUC. For example, many of the imputed variables were categorical such as job category, percent of job completed by a professional level employee, and the position of the majority worker on a job. The benefit of having an imputed data set that performs acceptably well is that each model is fed the full data set and therefore produces predictions for the full data. Beforehand, Random Forest and Logistic Regression models were limited to complete original data, which implied a reduced final data set. With complete sets of model predictions, the high performing models could be blended to test whether better predictions were possible. For this reason, the imputed data models were chosen to be progressed to the development of blended models.

@@

ggplot(auc.melt, aes(x= factor(L1), y= value)) + 
        geom_violin(aes(fill = greater, colour = greater)) + 
        geom_jitter(alpha = 0.2, shape = 16, colour= 'gray70') +
        stat_summary(fun.y="mean", geom="point", colour = "gray60", shape =3) +
        theme(axis.text.x=element_text(angle=45,hjust=1)) +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "AUC Distribution from 100 Blended Models of each Method") +
        scale_x_discrete(labels = c("boost" = "BT Blend", 
                         "log" = "FWLS", 
                         "average" = "Simple Average Blend",
                         "orig.log" = "Orignal LR", 
                         "rf" = "RF Blend",
                         "simp.boost" = "Simple BT Blend",
                         "simp.log" = "Simple LR Blend")) +
        scale_fill_manual(values = c('navajowhite','sienna1'),
                          name = "Mean AUC Greater than Original Logistic Regression") +
        scale_colour_manual(values = c('navajowhite','sienna1'),
                          name = "Mean AUC Greater than Original Logistic Regression") +
        theme(legend.position = "bottom")

@@

ggplot(auc.150, aes(x= factor(L1), y= value)) + 
        geom_violin(aes(fill = signif, colour = signif)) + 
        geom_jitter(alpha = 0.2, shape = 16, colour= 'gray70') +
        stat_summary(fun.y = "mean", geom="point", colour = "gray60", shape =3) +
        theme(axis.text.x=element_text(angle=45,hjust=1)) +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "AUC Distribution from 150 Blended Models of Best Methods") +
        scale_x_discrete(labels = c( 
                         "average" = "Simple Average Blend",
                         "orig.log" = "Orignal LR",
                         "simp.log" = "Simple LR Blend")) +
        scale_fill_manual(values = c('navajowhite','sienna1'),
                          name = "Mean AUC Significantly Greater than Original Logistic Regression") +
        scale_colour_manual(values = c('navajowhite','sienna1'),
                          name = "Mean AUC Significantly Greater than Original Logistic Regression") +
        theme(legend.position = "bottom")

@@

Trials of different methods of blending model results demonstrated that again, the simplest methods worked best with the data. In this case, averaging the results of the best three single models (Logistic Regression, Random Forests, and Boosted Trees) or taking a Logistic Regression of the three models outperformed more complex Boosted Trees, Random Forests, and Logistic Regression blends that facilitated interaction of the individual model results and original variables (meta-features). 
It was plausible that simple blending models would perform better than complex methods because the data was not big enough for complex models to learn intricate patterns. This was observed in the single method models (previous chapter) when Logistic Regression performed as well as Random Forests and Boosted Trees. The literature does not extensively cover blended machine learning methods however their success in the 2009 Netflix competition generated some publications. The Netflix data set comprised of almost 3,000,000 observations, so the size of the data could have enabled the success of complex blending methods such as FWLS [@Sill2009]. That being the case, it was still surprising that simple averaging of probabilities from the individual models performed almost as well as blending results using Logistic Regression. It was expected that Logistic Regression would have an advantage over model averaging due to its ability to weight the prediction variables while maintaining a simple model structure.

The Logistic Regression blending model did indeed weight predictions from the original models differently. A summary of the resulting coefficients from an example blended Logistic Regression model is shown in the table below.

@@

As shown in the table, if the predicted probability from the original Logistic Regression model increased by 0.1, the odds of the model predicting a loss making job increased by 56.0%. The weightings contrasted the other simple model that averaged the original probabilities (i.e. weighted them equally) but still managed to produce similar AUC results. It was not clear why this was the case, however both simple models were compared again in the analysis of the impact each model could have on the company's overall profits. The analysis of overall profits did not directly relate to AUC measurements, so a range of models were carried forward for comparison. These included the three original individual models, three simple blended models and three complex blended models.

@@

@@

The first research question, whether machine learning and statistical techniques can predict project profitability based on historic data, has been answered. The measure of profitability could not be modeled but simplifying the profitability prediction to a binary (profit/loss) model was successful. To answer the second research question, whether the model would improve the business' profitability, a profit analysis was performed.

The models output a 'probability' between 0 and 1 that each project will be a loss making job (probability = 1 indicates a loss making job). This leaves the question, at what probability would a decision-maker round the probability up to 1 or down to 0? And what business decision would then be made? To find the threshold point for rounding, an experimental business-scenario was tested. At a given threshold, all projects calculated above the threshold were considered too risky, and were rejected. All profits and losses from these projects were forfeited, while the profits from the remaining jobs (below the threshold) were summed to give a revised total profit.

@@

# How to achieve statistical power of 0.8?
finance.plot = readRDS("plot_finance.rds" )
fin.summary.1 = finance.plot %>% group_by(threshold, method) %>% 
        summarise_each(funs(mean_profit_ratio = round(mean(.), 1), 
                            sd_prof_ratio = round(sd(.), 1)),
                       profit.ratio)

@@

Using this logic, if the threshold was 0, all projects were rejected and the total profit would equal \$0. If the threshold was 1, all projects were accepted and the total profits would be the same as historical figures. This profit calculation was made on a range of thresholds between 0 and 1 at 0.05 increments. The total profits were plotted for each threshold and joined to make a profit curve where at an optimal threshold, the total profits were highest (refer plot below).

Profit curves for the three simple blended models as well as the three individual models were calculated. However, the total profit coordinates were based on probability outputs in test data sets. This means that the profit curve for the simple logistic regression blend (for example) would be different for different training/testing data divisions. I.e. the total profits at each point in the profit curve depend on the model parameters that were calculated using the training data and which projects were analysed in the test data. Therefore, the process of training a blended model, then calculating the profit curve from the test data was repeated multiple times to create a distribution of profit curves for each method (blended and individual).

To start, 10 curves for each method were calculated (2 x 5-fold cross validation). Two-sampled power calculations (for a power of 0.8) were performed between the results of each method at their optimal threshold against the best performing method (simple average blend). For example, the optimal threshold for the simple average blended model was 0.55 while the optimal threshold for the original Boosted Tree model was 0.60. The 10 'total profit' values at these two thresholds were compared in the two-sample power calculation to determine the number of samples required to achieve a power of 0.8. The maximum required number of samples was 64 so 100 models were of each method were built (20 x 5 fold cross validation).

The plot below illustrates the results of each model at each threshold with a line joining the mean value at each threshold for each method. The grey ribbon illustrates the 95% confidence interval for the profit improvement ratio at each threshold for the 100 models.

@@

Two blended models clearly outperformed the individual models as shown in the plot above with higher profit ratios as well as tighter confidence intervals. The table below denotes each models' optimal thresholds and the corresponding increase in total profits:

@@

The simple logistic regression blend performed best with the highest mean profit ratio at `r best.mean`% with a standard deviation of `r best.sd`%. This means that for the simple blended logistic regression model, if all jobs above the probability threshold `r best.t` were rejected, the profits would increase on average by `r best.mean - 100`% in comparison to historical profits.

For clarity, the plot for the simple blended logistic regression model is shown below:

@@

@@

### Highest Profit Curves

The simple average blend and logistic regression blended models outperformed the complex blended models, which is rational because their AUC values were significantly higher than the others. The simple average of Logistic Regression, Random Forests, and Boosted Trees results produced a profit curve with an almost identical profit improvement (and standard deviation) to the Logistic Regression that blended the same three variables. The blended Simple Logistic Regression produced a `r best.mean-100`% increase in profit by rejecting jobs with a probability of loss over `r best.t`. To put this in context, if the company's yearly profits were \$1,000,000, the mean yearly increased profit would be \$`r comma_format() ((best.mean-100)/100*1000000)` by rejecting all jobs with a probability higher than `r best.t`. In the data, projects assigned a probability higher than `r best.t` represent only `r ((simp.l %>% filter(mean.iter>best.t) %>% nrow)/nrow(simp.l)*100) %>% round(1)`% of all projects.

The plot below illustrates this proportion of projects as well as the proportion of profits and losses assigned to a range of probability levels by the simple logistic regression blend. The red line sits at its optimal threshold point, i.e. projects greater than the threshold point would be rejected:

reject.plot <- readRDS('reject_plot.rds')

ggplot(simp.l2, aes(x = prob.bin)) + 
        facet_grid(faceter~., scales = "free_y",
                   labeller = labeller(faceter = c("hist" = "Count of Projects",
                                       "profs"= "Relative Profits/Losses"
                                       ))) +
        geom_bar(data = simp.l2 %>% filter(faceter == "hist")) + 
        geom_bar(data = simp.l2 %>% filter(faceter == "profs"),
                 aes(x = prob.bin, y = profit.ratio), stat = "identity") +
        # geom_bar(data = simp.l2 %>% filter(faceter == "inv"),
                 # aes(x = prob.bin, y = invoiced.ratio), stat = "identity") +
        labs(x = "Probability Score", 
             y = "",
             title = "Distribution of Project Counts and Project Profits Against Probability") +
        geom_vline(xintercept = 10.5, colour = hcl(h = 15, l = 65, c = 100), linetype = 1,
                   alpha = 0.8) +
        theme(panel.background = element_rect(fill = 'gray95'))

@@

The tested profit curve models can be divided into three categories: original individual methods, simple blends of the individual methods, and complex blends of the individual methods that employed interactions with original features. All original methods did achieve improvements of profit at their optimal thresholds, but did not necessarily consistently improve profits across all cross-folded models. The shaded grey 95% confidence intervals in the Profit Curve plot shows that with the 100 sample models that were analysed, none of the lower bounds for the original methods' were above 100% on the y-axis. That is, it cannot be said with 95% confidence that the original methods' would produce a profit higher than historical profits if projects higher than the optimal thresholds (jobs that are likely to be loss making) were rejected. Clearly this is not helpful for the case-study business to adopt.

The complex blended models had mixed results. FWLS and Random Forests performed more poorly than the individual methods while Boosted Trees produced a marginally higher profitability ratio (`r best.fin %>% filter(method == "comp.boost") %>% .$mean_profit_ratio`) but the lower bound of its confidence interval was `r best.fin %>% filter(method == "comp.boost") %>% .$lower_bound_95CI`, i.e. not above original profits.

The simple blended models performed more favourably, as already discussed, particularly the simple average and Logistic Regression. The Boosted Tree's lower bound confidence interval was `r best.fin %>% filter(method == "simp.boost") %>% .$lower_bound_95CI`% at its optimal threshold point. This is not above 100%, and again does not make a convincing argument for the case-study to adopt it.

It is not clear why simple linear models and the averaging model outperformed ensemble tree blending methods. It may be that there was not enough data for the ensemble trees to adequately learn the complex series of rules they develop as previously stated. It was also surprising FWLS was ineffective given the success it had in the Netflix Prize competition [@Sill2009]. The only differences between FWLS and the simple Logistic Regression (that performed best) were two additional explanatory variables and four interaction terms. The number of variables was not high for the amount of data. According to @Peduzzi1996, 10 events per explanatory variable or more avoids the risk of biased estimation of variable coefficients in logistic regression. The data contained 315 events per explanatory variable and did not pose that risk. Nevertheless, the additional variables must have added misleading noise to the model. The performance of the models so far has been concerned with AUC and profit, however insight into the differences in model performance may be gained by reviewing confusion matrix statistics such as TPR and FPR.

>>>>

library('knitr')
library('ggplot2')
library("plyr")
library("dplyr")
library('magrittr')
library('reshape2')
library("rpart")
library('car')
library('e1071')
library('party')
library('randomForest')
library('RColorBrewer')
library('pwr')
library('scales')
library('purrr')

@@

A model that is perfectly classified would have all positives correctly classified (1.0 TPR) and no incorrectly classified positives (0.0 FPR). If a model has a 1.0 TPR and 1.0 FPR (top right corner) it has correctly classified all positives at the expense of incorrectly classifying all negatives as positive. It is 'dumbly' classifying all cases as positive. If a model classifies 0.8 of its positives as TP but also 0.8 of its negative cases as positive it is 'dumb' in a similar way to the previous example. There is an 80% chance *any* case will be classified as positive. Therefore, a point that lies on the diagonal line classifies cases with the same capability as random chance.

@@

For binary classification, AUC is a more meaningful statistic than classification accuracy when the output is a probability that can be applied to the problem. The AUC indicates model performance across many thresholds while classification accuracy represents a single threshold. The AUC statistic is therefore useful in problems where cases can be ranked and a cut-off threshold that will optimise the rates of true positives (TP's) and true negatives (TN's) can be found [@Huang2005]. Note the 'probability' outputs from the models are not true statistical probabilities, but a score the model has assigned to a case using its own measure of certainty.

@@

ggplot(reduc.totAUC, aes(x= factor(L1), y= value)) + 
        geom_violin(aes(fill = sigdiff, colour = sigdiff), alpha = 0.8) + 
        stat_summary(fun.y="mean", geom="point", colour = "gray30", shape =3) +
        stat_summary(aes(label = round(..y.., 3)), fun.y="mean",
                     geom="text", colour = "gray20", size =3, 
                     hjust = -.2, angle = 40, vjust = -.2
                     ) +
        scale_fill_manual(values = c('sienna1','navajowhite'),
                          name = "Significantly different from boosted trees") +
        scale_colour_manual(values = c('sienna1','navajowhite'),
                          name = "Significantly Different from Boosted Trees") +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "AUC Distribution from 100 models of Each Method") +
        scale_x_discrete(labels = c("boost" = "Boosted Trees", 
                         "log" = "Logistic Regression",
                         "naiive" = "Naive Bayes", "rf" = "Random Forest")) +
        theme(legend.position = "bottom")

@@

ggplot(mice.totAUC, aes(x= factor(L1), y= value)) + 
        geom_violin(aes(fill = sigdiff, colour = sigdiff), 
                    alpha = 0.8) + 
        stat_summary(fun.y="mean", geom="point", colour = "gray30", shape =3) +
        stat_summary(aes(label = round(..y.., 3)), fun.y="mean",
                     geom="text", colour = "gray20", size =3, 
                     hjust = -.2, angle = 40, vjust = -.2
                     ) +
        scale_fill_manual(values = c('sienna1','navajowhite'),
                          name = "Significantly different from boosted trees") +
        scale_colour_manual(values = c('sienna1','navajowhite'),
                          name = "Significantly different from boosted trees") +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "AUC Distribution from 100 Models of Each Method - Imputed Data") +
        scale_x_discrete(labels = c("boost" = "Boosted Trees", 
                         "log" = "Logistic Regression",
                         "naiive" = "Naive Bayes", "rf" = "Random Forest")) +
        theme(legend.position = "bottom")
```

@@

ggplot(combine.micer, aes(x= factor(L1), y= value)) +
        geom_violin(aes(fill = Data_Source, colour = Data_Source)) +
        # stat_summary(fun.y="mean", geom="point", colour = "gray60", shape =3) +
        scale_fill_manual(values = c('sienna1','navajowhite'),
                          name = "Data Source") +
        scale_colour_manual(values = c('sienna1','navajowhite'),
                          name = "Data Source") +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "Comparison of Original Data vs. Imputed Data") +
        scale_x_discrete(labels = c("boost" = "Boosted Trees", 
                         "log" = "Logistic Regression",
                         "naiive" = "Naive Bayes", "rf" = "Random Forest")) +
        theme(legend.position = "bottom")

@@

Boosted trees, Random Forests and Logistic Regression all had mean AUC values between 0.756 and 0.767. The tests were repeated using imputed data and again the means sat in a similar range from 0.755 to 0.764. Many variables were categorical which are not particularly well suited for imputation. These included variables such as job category, percent of job completed by a professional level employee, and the position of the majority worker on a job. Therefore it is logical the AUC's did not improve, in fact they marginally reduced. The benefit of having an imputed data set that performs acceptably well is that each model is fed the full data set and therefore produces predictions for the full data set. Before, Random Forest and Logistic Regression models were limited to complete original data, which implied a reduced final data set. With complete sets of model predictions, the high performing models could be blended to test whether better predictions were possible. For this reason, the imputed data models were chosen to be progressed to the development of blended models.

@@

ggplot(auc.melt, aes(x= factor(L1), y= value)) + 
        geom_violin(aes(fill = greater, colour = greater), 
                    alpha = 0.8) + 
        stat_summary(fun.y="mean", geom="point", colour = "gray30", 
                     shape =4) +
        stat_summary(aes(label = round(..y.., 3)), fun.y="mean",
                     geom="text", colour = "gray20", size =3, 
                     hjust = -.2, angle = 40, vjust = -.2
                     ) +
        theme(axis.text.x=element_text(angle=40,hjust=1)) +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "AUC Distribution from 100 Blended Models of each Method") +
        scale_x_discrete(labels = c("boost" = "BT Blend", 
                         "log" = "FWLS", 
                         "average" = "Simple Average Blend",
                         "orig.log" = "Orignal LR", 
                         "rf" = "RF Blend",
                         "simp.boost" = "Simple BT Blend",
                         "simp.log" = "Simple LR Blend")) +
        scale_fill_manual(values = c('navajowhite','sienna1'),
                          name = "AUC Greater than Original Logistic Regression") +
        scale_colour_manual(values = c('navajowhite','sienna1'),
                          name = "AUC Greater than Original Logistic Regression") +
        theme(legend.position = "bottom")

@@

ggplot(auc.150, aes(x= factor(L1), y= value)) + 
        geom_violin(aes(fill = signif, colour = signif),
                    alpha = .8) + 
        stat_summary(fun.y = "mean", geom="point", 
                     colour = "gray30", shape =3) +
        stat_summary(aes(label = round(..y.., 3)), fun.y="mean",
                     geom="text", colour = "gray20", size =3, 
                     hjust = -.2, angle = 40, vjust = -.2
                     ) +
        theme(axis.text.x=element_text(angle=40,hjust=1)) +
        labs(x = "Method", y = "Area Under Curve (AUC)",
             title = "AUC Distribution from 150 Blended Models of Best Methods") +
        scale_x_discrete(labels = c( 
                         "average" = "Simple Average Blend",
                         "orig.log" = "Orignal LR",
                         "simp.log" = "Simple LR Blend")) +
        scale_fill_manual(values = c('navajowhite','sienna1'),
                          name = "AUC Significantly Greater than Original Logistic Regression") +
        scale_colour_manual(values = c('navajowhite','sienna1'),
                          name = "AUC Significantly Greater than Original Logistic Regression") +
        theme(legend.position = "bottom")

@@

```{r blend_vs_initial}
auc.150 = readRDS("report_data/blend_auc150.rds")
mice.totAUC <- readRDS("report_data/mice_df_auc.rds")
compare.blends <- append(auc.150, mice.totAUC)

index = c(3,4,5,6)
pvals = data.frame("method" = names(compare.blends)[index], 
                   "mean_AUC" = map_dbl(compare.blends, function(x) round(mean(x),3))[index],
                   'p_value' = NA)

for(i in seq_along(index)){
        pvals$p_value[i] = t.test(compare.blends[[3]],
                                  compare.blends[[index[i]]])$p.value %>%
                format(digits = 2) %>% as.numeric
}

pvals[1,'p_value'] <- NA
pvals <- pvals %>% arrange(desc(mean_AUC))

new.levels = c("simp.log" = "SBLR",
               "log" = "Original Logistic Regression",
               "boost" = "Original Boosted Trees",
               "rf" = "Original Random Forest")

pvals <- mutate(pvals, method = revalue(method, new.levels))
names(pvals) = c("Method", "Mean AUC", "P-Value for t-test against SBLR")

The AUC distributions from the two best blended models, simple Logistic Regression and simple averaging, were statistically significantly higher than the AUC distribution of the original Logistic Regression model. Blended models improved the mean AUC from 0.759 to 0.77, which is an increase of `r ((0.77-0.759)/0.759) %>% "*"(100) %>% round(1)`%. The original Logistic Regression model was chosen for comparison because it was a fast baseline model that had an AUC distribution not statistically different to the other high performing original models. It was expected that model blending would improve predictions because it combined the strengths of individual models with different theoretical foundations. Next, the performance of different blended models is discussed.

The trials of different blending methods demonstrated that again, the simplest methods worked best with the case study data. In this incident, averaging the results of the best three single models (Logistic Regression, Random Forests, and Boosted Trees) or taking a Logistic Regression of the three models outperformed more complex Boosted Trees, Random Forests, and Logistic Regression blends that facilitated interaction of the individual model results and original variables (meta-features).

Simple blended models could perform better than complex methods if the data is not big enough for the complex models to learn the more intricate patterns at which they excel. This was observed in the single model methods (previous chapter) when Logistic Regression performed as well as Random Forests and Boosted Trees. There is limited literature on blended machine learning methods, however their success in the 2009 Netflix competition generated some publications. The Netflix data set comprised of almost 3,000,000 observations, so the size of the data could have enabled the success of complex blending methods such as FWLS [@Sill2009]. That being the case, it was still surprising that simple averaging of probabilities from the individual models performed almost as well as Logistic Regression. Logistic Regression's ability to weight the prediction variables while maintaining a simple model structure was expected to give an advantage over averaging.

The blended Logistic Regression model did indeed weight predictions from the original models differently. A summary of the resulting coefficients from an example blended Logistic Regression model is shown in the table below.

@@

As displayed in the table, if the predicted probability from the original Logistic Regression model increased by 0.1, the odds of the blended model predicting a loss making job increased by 56.0%. This same interpretation can be made for the Random Forest and Boosted Trees values, which means the results from Boosted Tree models were weighted highest. Clearly, the weightings contrasted the equal weightings from the simple averaged model, yet their AUC results were similar. It was not clear why this was the case, however both simple models were compared again in the analysis of the impact each model could have on the company's overall profits. The analysis of overall profits did not directly relate to AUC measurements, so a range of models were carried forward for comparison. These included the three original individual models, three simple blended models and three complex blended models.

@@

This chapter first presents the full range of method results in terms of improvements to the case study's bottom line. A business decision making scenario was created so that profit curves based on the decision rule could be built and analysed. From the profit curves, optimal probability thresholds could be derived for each method (individual and blended methods). These threshold points allowed the probability output from each model to be converted to categorical output (profit or loss) which in turn could generate confusion matrix statistics for each method. The confusion matrix statistics are discussed after the profit curve results and discussion. Then, output from the method which produced the best profit curve was used to create two subsets of profitable and loss mkaing projects. The characteristics of these subsets are compared and contrasted to gauge any overall trends. Finally, alternate ways of framing the business decision scenario are proposed.

@@

The first research question, whether machine learning and statistical techniques can predict project profitability based on historic data, has been answered. The measure of profitability could not be modeled well, but simplifying the profitability prediction to a binary (profit/loss) model was successful and a mean AUC of 0.77 was achieved for the best blended model. This falls midway between a model that predicts perfectly (AUC = 1) and a model that predicts as well as random chance (AUC = 0.5). To answer the second research question, whether the model would improve the business' profitability, a profit analysis was performed.

The predictive models output a 'probability' between 0 and 1 that each project will be a loss making job (where probability = 1 indicates a loss making job). The question the arises, at what probability would a decision-maker round the probability up to 1 or down to 0? And what business decision would then be made? To find the threshold point for rounding, an experimental business-scenario was tested. At a given threshold, all projects with probability outputs above the threshold were considered too risky, and were rejected. All profits and losses from these projects were forfeited, while the profits from the remaining jobs (below the threshold) were summed to give a revised total profit.

@@

# How to achieve statistical power of 0.8?
finance.plot = readRDS("report_data/plot_finance.rds" )
fin.summary.1 = finance.plot %>% group_by(threshold, method) %>% 
        summarise_each(funs(mean_profit_ratio = round(mean(.), 1), 
                            sd_prof_ratio = round(sd(.), 1)),
                       profit.ratio)

@@

Using this logic, if the threshold was 0, all projects were rejected and the total profit would equal \$0. If the threshold was 1, all projects were accepted and the total profits would be the same as historical figures. This profit calculation was made on a range of thresholds between 0 and 1 at 0.05 increments. The total profits were plotted for each threshold and joined to make a profit curve where at the optimal threshold, the total profits were highest (refer plot below).

Profit curves for the three simple blended models as well as the three individual models were calculated. However, the total profit coordinates were based on probability outputs in test data sets. This means that the profit curve for the simple Logistic Regression blend (for example) would be different for different training/testing data divisions. In other words, the total profits at each point in the profit curve depend on the model parameters that were calculated using the training data and which projects were analysed in the test data. Therefore, the process of training a blended model, then calculating the profit curve from the test data, was repeated multiple times to create a distribution of profit curves for each method (blended and individual).

To start, 10 curves for each method were calculated (using 2 x 5-fold cross validation). Two-sampled power calculations (for a power of 0.8) were performed between the results of each method (at their optimal threshold points) against the best performing method (simple Logistic Regression blend). For example, the optimal threshold for the simple Logistic Regression blended model was 0.55 while the optimal threshold for the original Boosted Tree model was 0.60. The 10 'total profit' values at these two thresholds were compared in the two-sample power calculation to determine the number of samples required to achieve a power of 0.8. Out of all the methods compared against the Blended Logistic Regression, the maximum required number of samples was 64, so 100 models of each method were built (20 x 5-fold cross validation).

The plot below illustrates the distrubtion of profit curves for each method, where the solid lines join the mean values at each threshold point. The grey ribbon illustrates the 95% confidence interval for the profit improvement ratio at each threshold point for the 100 models.

@@

Two blended models clearly outperformed the individual models as shown in the above plot with higher profit ratios as well as tighter confidence intervals. The table below denotes each models' optimal thresholds and the corresponding increase in total profits:

@@

The simple Logistic Regression blend performed best with the highest mean profit ratio of `r best.mean`% with a standard deviation of `r best.sd`%. This means that for the simple blended Logistic Regression model, if all jobs above the probability threshold `r best.t` were rejected, the profits would increase on average by `r best.mean - 100`% in comparison to historical profits.

For clarity, the plot for the simple blended Logistic Regression model is shown below:

@@

```
<!-- comments on your results; -->
<!-- explains what your results mean; -->
<!-- interprets your results in a wider context; indicates which results were expected or unexpected; -->
<!-- provides explanations for unexpected results. -->
<!-- The Discussion should also relate your specific results to previous research or theory. You should point out what the limitations were of your study, and note any questions that remain unanswered. The Discussion CAN also include Conclusions/Future Research. -->

In this section, the results from the models that delivered the highest performing profit curves are reviewed first and analysed in a broader business context. Then, all the methods that produced profit curves are compared and possible explanations for differences are evaluated.

### Examination of the Highest Profit Curves

Profit curves from the simple average blend and Logistic Regression blended models outperformed the complex blended models, which follows logically from their significantly higher AUC distributions. The simple average blended model produced a profit curve with an almost identical profit improvement (and standard deviation) to the Logistic Regression blend. As summarised in the previous section, the blended Simple Logistic Regression produced a `r best.mean-100`% increase in profit by rejecting jobs with a probability of loss over `r best.t`. In a business context, this means that if the company's yearly profits were \$1,000,000, the mean increased profit could be \$`r comma_format() ((best.mean-100)/100*1000000)` by rejecting all jobs with a probability higher than `r best.t`. In the data, projects assigned a probability higher than `r best.t` represent only `r ((simp.l %>% filter(mean.iter>best.t) %>% nrow)/nrow(simp.l)*100) %>% round(1)`% of all projects.

The plot below illustrates the relative profits and losses that sit within the range of probability levels, as assigned by the simple Logistic Regression blend.

$Relative\ Profit_t = \frac{Profit_t}{\sum_{t = 1}^{n}\left|Profit_t\right|}$

$$
\begin{aligned}
Profit_t &= \textrm{Profit from project}\ t\textrm{. Negative values indicate loss making projects}
\end{aligned}
$$

The red line sits at the profit curve's optimal threshold point, 0.6, so projects greater than the threshold point would be rejected:

reject.plot <- readRDS('report_data/reject_plot.rds')

ggplot(simp.l2, aes(x = prob.bin)) + 
        facet_grid(faceter~., scales = "free_y",
                   labeller = labeller(faceter = c("hist" = "Count of Projects",
                                       "profs"= "Relative Profits and Losses"
                                       ))) +
        geom_bar(data = simp.l2 %>% filter(faceter == "hist")) + 
        geom_bar(data = simp.l2 %>% filter(faceter == "profs"),
                 aes(x = prob.bin, y = profit.ratio), stat = "identity") +
        # geom_bar(data = simp.l2 %>% filter(faceter == "inv"),
                 # aes(x = prob.bin, y = invoiced.ratio), stat = "identity") +
        labs(x = "Probability Score", 
             y = "",
             title = "Distribution of Project Counts and Project Profits Against Probability") +
        geom_vline(xintercept = 10.5, 
                   colour = hcl(h = 15, l = 65, c = 100), linetype = 1,
                   alpha = 0.8) +
        geom_hline(yintercept = 0, colour = "white", linetype = 1)
        theme(panel.background = element_rect(fill = 'gray95'))

@@

As shown in the upper plot, a low count of projects fall above the optimal threshold for the simple Logistic Regression blend. The lower plot illustrates the relative profits (and losses for negative values) within each probability score bin which are normalised by the sum of the absolute value of all profits and losses. The model has assigned low probability scores to profitable projects and high scores to unprofitable jobs well. The heights of the bars reduce from left to right, and loss making jobs dominate projects on the right hand side of the optimal threshold point. This is advantageous as these jobs would be rejected.

@@

The tested profit curve models can be divided into three categories: original individual methods, simple blends of the individual methods, and complex blends of the individual methods that employed interactions with original features. All original methods did achieve improvements of profit at their optimal thresholds, but did not necessarily consistently improve profits across all cross-folded models. The shaded grey 95% confidence intervals in the Profit Curve plot shows that with the 100 sample models that were analysed, none of the lower bounds for the original methods' were above 100% on the y-axis. That is, it cannot be said with 95% confidence that the original methods' would produce a mean profit higher than historical profits in the given decision framework. This framework stated that projects higher than the optimal thresholds (jobs that are likely to be loss making) should be rejected. Clearly this level of certainty is not beneficial for the case-study business to adopt.

The complex blended models had mixed results. FWLS and Random Forests performed more poorly than the individual methods while Boosted Trees produced a marginally higher profitability ratio (`r best.fin %>% filter(method == "comp.boost") %>% .$mean_profit_ratio`%) but the lower bound of its confidence interval was `r best.fin %>% filter(method == "comp.boost") %>% .$lower_bound_95CI`%, i.e. not an improvement on original profits.

The simple blended models performed much more favourably, as already discussed, particularly the simple average and Logistic Regression blends. The blended Boosted Tree's lower bound confidence interval was `r best.fin %>% filter(method == "simp.boost") %>% .$lower_bound_95CI`% at its optimal threshold point. This is not above 100%, and again does not make a convincing argument to the case study business.

It is not clear why simple linear models and the averaging model outperformed ensemble tree blending methods. As previously stated, the ensemble trees may not have received enough data to adequately learn the complex series of rules they develop. It was also surprising FWLS was ineffective given the success it had in the Netflix Prize competition [@Sill2009]. The only differences between FWLS and the simple Logistic Regression (that performed best) were two additional explanatory variables and four interaction terms. The number of variables was not high for the amount of data since according to @Peduzzi1996, 10 events per explanatory variable or more avoids the risk of biased estimation of variable coefficients in Logistic Regression. The data contained 315 events per explanatory variable and did not pose that risk. Nevertheless, the additional variables must have added misleading noise to the model.
